Enhancing Adversarial Robustness for Deep Metric Learning

### 2021-10-25 robdml.tex
tex

Then, \emph{Hardness Manipulation} (HM) aims to increase the \emph{source}
hardness $H_\mathsf{S}$ into a specified ``\emph{destination} hardness''
$H_\mathsf{D}$, by finding adversarial examples of the source triplet $(\mA,
\mP, \mN)$, \ie, $(\mA + {\vr}_a, \mP + {\vr}_p, \mN + {\vr}_n)$, where
$({\vr}_a, {\vr}_p, {\vr}_n)$ are the adversarial perturbations.
%
If we denote the hardness of the adversarially perturbed \emph{source} triplet
as $\tilde{H}_\mathsf{S}$, the HM can be implemented as follows:
%
\begin{equation}
	%
	\hat{\vr}_a, \hat{\vr}_p, \hat{\vr}_n = \argmin_{\vr_a,\vr_p,\vr_n}
	\Big\|\max(0, H_\mathsf{D} - \tilde{H}_\mathsf{S} ) \Big\|_2^2.
	%
	\label{eq:hm}
	%
\end{equation}
%
The $\max(0,\cdot)$ part in \cref{eq:hm} truncates the gradient when
$H_\mathsf{S}>H_\mathsf{D}$, automatically stopping the optimization, because
$H_\mathsf{S}$ is not desired to be reduced once being greater than
$H_\mathsf{D}$.
%
The \cref{eq:hm} is written in the L-$2$ norm form instead of the standard Mean Squared
Error because HM can be directly extended into vector form for a
mini-batch.
%
The optimization problem can be solved by projected gradient
descent~\cite{madry}.
%
And the resulting adversarial examples are used for adversarially training the
DML model:
%
$L_\text{T}(\phi(X_a+\bar{r}_a), \phi(X_p^S+\bar{r}_p),
\phi(X_n^S+\bar{r}_n))$.

\begin{align}
	&
	\sign\big\{
		-\frac{\partial}{\partial r} \big\| \min(0, 
		\tilde{H}_\mathsf{S} - H_\mathsf{D}) \big\|_2^2
	\big\}
	\\
	=&
	\sign\big\{
		-2(\tilde{H}_\mathsf{S}-H_\mathsf{D})
		\frac{\partial }{\partial r} \tilde{H}_\mathsf{S} \big\}
	=
	\sign\big\{
		\frac{\partial}{\partial r} \tilde{H}_\mathsf{S}
		\big\}
\end{align}

@@

Instead of directly maximizing the source hardness, we propose to stop the
maximization process when the adversarial examples reach a certain level of
hardness in order to avoid model collapse.
%
Based on the same anchor image $X_a$, we sample another positive image $X_p^D$
and negative image $X_n^D$ with a different triplet sampling strategy, and
we call its hardness as ``destination hardness'', denoted as $H_\mathsf{D}$.

>>>>

Then, \emph{Hardness Manipulation} (HM) aims to increase the \emph{source}
hardness $H_\mathsf{S}$ into a specified ``\emph{destination} hardness''
$H_\mathsf{D}$, by finding adversarial examples of the source triplet $(\mA,
\mP, \mN)$, \ie, $(\mA + {\vr}_a, \mP + {\vr}_p, \mN + {\vr}_n)$, where
$({\vr}_a, {\vr}_p, {\vr}_n)$ are the adversarial perturbations.
%
If we denote the hardness of the adversarially perturbed \emph{source} triplet
as $\tilde{H}_\mathsf{S}$, the HM can be implemented as follows:
%
\begin{equation}
	%
	\hat{\vr}_a, \hat{\vr}_p, \hat{\vr}_n = \argmin_{\vr_a,\vr_p,\vr_n}
	\big\|\max(0, H_\mathsf{D} - \tilde{H}_\mathsf{S} ) \big\|_2^2.
	%
	\label{eq:hm}
	%
\end{equation}
%
The $\max(0,\cdot)$ part in \cref{eq:hm} truncates the gradient when
$H_\mathsf{S}>H_\mathsf{D}$, automatically stopping the optimization, because
$H_\mathsf{S}$ is not desired to be reduced once being greater than
$H_\mathsf{D}$.
%
The \cref{eq:hm} is written in the L-$2$ norm form instead of the standard Mean Squared
Error because HM can be directly extended into vector form for a
mini-batch.
%
The optimization problem can be solved by Projected Gradient
Descent (PGD)~\cite{madry}.
%
And the resulting adversarial examples are used for adversarially training the
DML model:
%
$L_\text{T}(\phi(\mA+\hat{\vr}_a), \phi(\mP+\hat{\vr}_p),
\phi(\mN+\hat{\vr}_n))$.
%
We abbreviate the adversarial training with adversarial examples created through
this way as $\text{HM}[H_\mathsf{S},H_\mathsf{D}]$ in the following text.

% strong adversary

Note, in the PGD case, the sign of negative gradient of the HM objective with
respect to an adversarial perturbation $\vr$ is equivalent to the sign of
gradient for directly maximizing the hardness $\tilde{H}_\mathsf{S}$ (hence
maximizing $L_\text{T}$), \ie,
%
\begin{align}
	\Delta \vr
	=&
	\sign\big\{
		-\frac{\partial}{\partial \vr} \big\| \max(0, 
		H_\mathsf{D} - \tilde{H}_\mathsf{S} ) \big\|_2^2
	\big\}
	\\
	=&
	\sign\big\{
		2(H_\mathsf{D} - \tilde{H}_\mathsf{S})
		\frac{\partial }{\partial \vr} \tilde{H}_\mathsf{S} \big\}
	=
	\sign\big\{
		\frac{\partial}{\partial \vr} \tilde{H}_\mathsf{S}
		\big\},
\end{align}
%
assuming that $H_\mathsf{D}>\tilde{H}_\mathsf{S}$.
%
And the perturbation $\vr$ is updated as $\vr \leftarrow \vr + \alpha \Delta \vr$
by PGD, where $\alpha$ is a step size parameter.
%
Thus, the optimization of HM objective can be interpreted as directly
maximizing $\tilde{H}_\mathsf{S}$, but stops very early once it reaches
$H_\mathsf{D}$.
%
This means the model learns from an efficient adversary during adversarial
training with HM.

Instead of directly maximizing the source hardness, we propose to stop the
maximization process when the adversarial examples reach a certain level of
hardness in order to avoid model collapse.

@@

%
Based on the same anchor image $X_a$, we sample another positive image $X_p^D$
and negative image $X_n^D$ with a different triplet sampling strategy, and
we call its hardness as ``destination hardness'', denoted as $H_\mathsf{D}$.
